An Analysis of Performance Bottlenecks in MRI Pre-Processing

### 2024-04-17 paper/paper.tex
[FIX] comments from Tristan.

\usepackage{cite}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=black,citecolor=blue,filecolor=black,urlcolor=blue}
\graphicspath{{figures}}
\usepackage{caption}
\usepackage{subcaption}
\captionsetup[subfigure]{singlelinecheck=false, skip=0pt}
\usepackage{listings}
\renewcommand{\lstlistingname}{Algorithm}
\usepackage[binary-units=true]{siunitx}
\usepackage{csvsimple}
\usepackage{booktabs}
\usepackage{amsmath}

@@

\begin{abstract}
	Magnetic Resonance Image (MRI) pre-processing is a critical step for neuroimaging analysis. However, the computational cost of MRI pre-processing pipelines is a major bottleneck for large cohort studies and some clinical applications. While High-Performance Computing (HPC) and, more recently, Deep Learning have been adopted to accelerate the computations, these techniques require costly hardware and are not accessible to all researchers. Therefore, it is critical to understand the performance bottlenecks of MRI pre-processing pipelines to improve their performance. Using Intel VTune profiler, we characterize the bottlenecks of several commonly used MRI-preprocessing pipelines from the ANTs, FSL, and FreeSurfer toolboxes. We found that few functions contribute to most of the CPU time, and that linear interpolation is the largest contributor. Data access is also  a substantial bottleneck. Last, we identified a bug in the ITK library that impact the performance of ANTs pipeline in single-precision and a potential issue with the OpenMP scaling in FreeSurfer recon-all. Our results provide a reference for future efforts to optimize MRI pre-processing workflows.
\end{abstract}

@@

% - No comprehensive study of the performance bottleneck of MRI pre-processing.
\HL{To the best of our knowledge, this is the first study that} characterize the computational cost of several commonly adopted MRI pre-processing pipelines. The results of our analysis serve as a reference for future efforts to optimize MRI pre-processing workflows.

@@

% - Pipelines
Neuroscientists developed various toolboxes to tackle the challenging task of pre-processing MRI data. We focus on the commonly accepted fMRIPrep~\cite{Esteban2019-bl} workflow as it a comprehensive pre-processing pipeline for both sMRI and fMRI. fMRIPrep combines several pipelines to produce a complete pre-processing workflow. We also study its sub-workflow separately to provide a finer grain analysis of the performance bottleneck.

% - data: OpenNeuro ds004513
% We use the OpenNeuro ds004513 v1.0.2 dataset~\cite{ds004513:1.0.2}. We chose this dataset for its wide range of age and equal distribution of biological sex. Using a diverse dataset should help to prevent findings of potential performance bottleneck associated to a specific population. \TG{Move this to methods}

% - VTune profiler
To profile the pipelines, we use the Intel VTune profiler~\cite{vtune-profiler}. The VTune profiler is a multi-language profiler with low performance overhead and provides information of functions at runtime level. We profile the application using \HL{-- single-threaded and 32 threads to simplify the analysis and study the the multi-threading performance, repsectively}. We aggregate subjects' results for each pipeline to obtain an average performance profile. We compiled each application with debug information to obtain human readable information on function and module names in VTune.

\section{Materials \& Methods}
\subsection{Pipelines}
We study the performance bottleneck of fMRIPrep; a commonly used fMRI and sMRI pre-processing workflow. First, we profile the entire workflow to obtain a coarse view of the workflow's performance. Furthermore, to characterize the performance at a finer grain, we profile the sub-workflows of fMRIPrep: ANTs brainExtraction, ANTs registrationSyN~\cite{Avants2008-ea}, FreeSurfer recon-all~\cite{Dale1999-wu}, FSL FAST~\cite{Zhang2001-hx}, FSL FLIRT~\cite{Jenkinson2002-od,Jenkinson2001-eu,Greve2009-dw}, and FSL MCFLIRT~\cite{Jenkinson2002-od}. We omitted some sub-workflows of fMRIPrep as they were not compatible with our dataset \TG{add a few words to explain why}. \TG{Information is missing to explain what is the statistical analysis in fMRIPrep. Likewise, it's unclear what ANTS registration, FSL Flirt, etc are and how they are combined in fMRIPrep. This section should start with a higher-level overview of fMRIPrep. You should also explain that fMRIPrep's building blocks are widely used in neuroimaging.} Moreover, the statistical analysis of the fMRIPrep workflow is not computationally expensive and is not the focus of this paper.

We use the default parameters for fMRIPrep and most of the sub-workflows. However, for FreeSurfer recon-all, we mimic the fMRIPrep specific 3 stages workflow \TG{This is cryptic, it needs more context.}.

We profile both the single and double precision version for ANTs registrationSyN and ANTs brainExtraction. ANTs uses ITK templates to implement the single precision alternative. \TG{You should re-order this paragraph as follows: (1) ANTs registration has two implementations, a double precision one and a single precision one. (2) This is actually coming from ITK. (3) We benchmark both implementations.}

\subsection{Data}
To ensure that our performance profiles are inclusive of different populations, we used data with a wide range of age and equal distribution of sex. We used the the OpenNeuro ds004513 v1.0.2 dataset~\cite{ds004513:1.0.2}. The anatomical, functional, and diffusion data from 20 health individuals was acquired from two different cohorts. The within-subject cohort has nine participants (mean age=43~yrs, std=7~yrs; 4 females) with two session: eyes open and eyes closed. The replication cohort has eleven participants (mean age=27~yrs, std=5~yrs; 6 females) with only the eyes open session. We profiled the application using both cohorts and the eyes open session, unless explicitly specified \TG{is it ``explicitly specified'' anywhere? If not, remove this.}.

\subsection{Profiling}
Profiling MRI pre-processing pipelines \TG{the terms pipelines, workflfows, applications are used interchangeably, let's harmonize and use only 1 term (I don't mind which one)} raises a few challenges. Neuroimaging pipelines use several programming languages including \textit{C}, \textit{C++}, \textit{Fortran}, \textit{Matlab}, and \textit{Python}, with several pipelines utilizing a combination of multiple languages. Pipelines are generally complex and computationally expensive.

The Intel VTune profiler addresses these challenges by offering multi-language support, low performance overhead, and information of functions at runtime level. However, common profiling challenges remain. Profilers require source code to be compiled with debug symbols to report interpretable information about the different functions and modules. Moreover, different input data can substantially impact the performance results due to conditional branching in the application and convergence thresholds. We discuss our approach to these challenges in the next section.
\MD{TODO Cite effect from varying input data on numerical stability and reduced precision. Ask yohan for references}

First, to obtain human readable information on function and module names in VTune, we re-compiled each pipeline with debug information in a Docker image. For use on HPC systems, we created Apptainer~\cite{Kurtzer2017-bu} images using Docker2Singularity\footnote{\href{https://github.com/singularityhub/docker2singularity}{https://github.com/singularityhub/docker2singularity}} \lstlistingname~\ref{alg:vtune_example} \TG{Could this be called a listing rather than an algorithm?} illustrates the profiling of a pipeline by mounting the VTune binary to the Apptainer image during execution.

\begin{minipage}{\linewidth}
	\lstinputlisting[
		language={sh},
		caption={Profiling an application within a Apptainer container, using VTune profiler.},
		captionpos=b,
		label=alg:vtune_example,
		numbers=left,
		frame=single,
		basicstyle=\footnotesize,
		numbersep=5pt,
		numberstyle=\tiny\color{gray},
		rulecolor=\color{black},
		tabsize=2,
	]{vtune-example.sh}
\end{minipage}

Second, we aggregate the profiling results of each pipeline across all subjects to obtain an average performance profile. The Pareto principle is a power law which states that 80\% of the effects come from 20\% of the causes. We hypothesize that the Pareto principle applies to the performance of MRI pre-processing pipelines, which is consistent with the literature~\cite{Kukunas2015-jd}. Therefore, for fine-grained analysis, we only study the functions with the largest contribution to CPU time, up to 80\% cumulative total runtime of the application \TG{This discussion on Pareto principle should be moved to the results section}. This heuristic allows us to focus our efforts on the performance critical sections of a pipeline. We report the mean and standard deviation of the execution time for those functions.

Last, we profile the application using two threading configuration: single-threaded and 32 threads. The single-threaded approach simplifies the analysis and limits the potential overhead from I/O on the profiling. 
The 32-thread approach uses all available CPU cores from a compute node of our infrastructure to study the multi-threading performance of the application. This is a more realistic usage for the user of the application.

Before profiling each application, we transfer the input data from the shared file system to the compute node. After profiling, we transfer back the output to the shared file system. This limits profiling variability due to I/O contingency.

The data generated by the VTune profiler is written to a shared file system, for administrative reasons. In principle, this could increase the profiling overhead. However, in practice there is no impact since the data generated by the VTune profiler is very small and is not written on the same file system as data produced by the applications.

\subsection{Metrics Definition}
VTune provides several metrics to characterize the performance of an application.  The makespan is the elapsed time from the start to the end of the application. The CPU time refers to the total time spent by the CPU, across all cores, to execute the application.

\begin{figure*}[ht]
	\noindent
	\begin{minipage}
		{0.5\textwidth}
		\begin{align*}
			C & = CPU\_CLK\_UNHALTED.THREAD                \\
			S & = CYCLE\_ACTIVITY.STALLS\_LDM\_PENDING     \\
			O & = CYCLE\_ACTIVITY.STALLS\_L1D\_PENDING     \\
			T & = CYCLE\_ACTIVITY.STALLS\_L2\_PENDING      \\
			W & = MEM\_L3\_WEIGHT                          \\
			H & = MEM\_LOAD\_UOPS\_RETIRED.LLC\_HIT        \\
			R & = MEM\_LOAD\_UOPS\_MISC\_RETIRED.LLC\_MISS \\
			M & = \frac{W \times R}{H + R}                 \\
			N & = \frac{H}{H + W \times R}                 
		\end{align*}
	\end{minipage}
	\begin{minipage}
		{0.5\textwidth}
		\begin{align}
			\%~Memory~Bound & = \frac{S}{C} \times 100          \\
			\%~L1~Bound     & = \frac{S - O}{C} \times 100      \\
			\%~L2~Bound     & = \frac{O - T}{C} \times 100      \\
			\%~L3~Bound     & = \frac{T \times N}{C} \times 100 \\
			\%~DRAM~Bound   & = \frac{T \times M}{C} \times 100 
		\end{align}
	\end{minipage}
	\caption{Intel performance monitoring events and derived memory metrics}
	\label{fig:memory-metrics}
\end{figure*}	
The memory bound metrics counts the number of cycles where the CPU is starved while waiting for in-flight memory demand load. L1 bound occurs when a demand load stall some CPU cores although the data already reside in L1 cache. Similarly, L2, L3, and DRAM bound occurs when a demand load stall the CPU and there is a cache miss in the previous level of cache, but the data reside in the L2, L3, and DRAM respectively. The store bound occurs when a store operation stall the CPU. Figure~\ref{fig:memory-metrics} shows the monitoring events and equation to derive the metrics defining the memory bound characterization~\cite{Intel2006-lc} \TG{This figure should go to appendix, it's not critical to the understanding of the paper}. We use the convention from~\cite{Kukunas2015-jd} for the equations.

@@

\subsection{CPU time distribution across functions was long tailed}
Figure~\ref{fig:long-tail-distribution} shows that 80\% of the total CPU time was spent in less than 1.5\% of the function across all pipelines. While consistent with the Pareto principle, the distribution of the functions' CPU time does not follow a Pareto distribution. We plotted the distribution of the functions' CPU time on a log-log scale and found it to be closer \TG{``closer'' is too vague: is it well approximated by a Zipf's law? Can you provide a p-value of this claim?} to a Zipf's law distribution with exponent $\alpha=1.8$. \TG{Add a few sentences to explain why this is interesting.}

@@

\subsection{Makespan: A performance overview}
Figure~\ref{fig:makespan-1thread} depicts the total time elapsed from the start to the end of each pipeline; i.e. makespan. There is a significant difference between the makespan of different applications. This is expected as the pipelines perform different tasks. The standard deviation is nearly null for all pipelines, except for FreeSurfer recon-all which as a small variation. Overall, this shows that the input data does not substantially affect the makespan of those pipelines. \TG{not sure what to make from this result, I don't find it particularly noticeable, I think I would omit it.}

\begin{figure}[t]
	\centering
	\includegraphics[width=\linewidth]{figures/makespan-1thread.png}
	\caption{Average makespan of the pipeline while using 1-thread. The error bars show the standard deviation. \TG{Font size in the figure is too small, it should be similar to the caption font size}}
	\label{fig:makespan-1thread}
\end{figure}

\subsection{Impact of Memory Bounded Functions}
Table~\ref{extab:memory-bound} shows that fetching data from disk is the main source of cycles where the CPU is starved \TG{I don't see disk in table 1}. Following, the L1 bound has the highest cache-level contribution to the memory bound of the pipelines. ANTS pipelines have the highest L1 bound, followed by FreeSurfer recon-all and FSL MCFLIRT. L2 and L3 bound is low for all pipeline, although highest for FreeSurfer recon-all. The DRAM bound is high for both FSL FLIRT and FreeSurfer recon-all, followed by ANTS brainExtraction in single and double precision. Store bound is low for all pipelines. We note an interesting relationship between the low L1 bound and the high DRAM bound of FreeSurfer recon-all and FSL FLIRT, compared to the other pipelines. \TG{Add discussion: what is interesting about this result?}

@@

\subsection{Main Bottleneck: Linear Interpolation}
Table~\ref{extab:interpolation} shows that interpolation (or convolution) is a critical part for all pipelines, except FreeSurfer recon-all. For those pipelines, interpolation contributes between 32.20\% to 62.70\% of the total CPU time. Overall, the number of functions using interpolation is low with less than 20 in each applications. In other words, few interpolation functions are contribute to a large portion of the CPU time. \TG{This should be moved up in the results, I think right after Figure 2, as this is one of the most interesting results.} \TG{Add a few sentences to explain why this is interesting.}
% There seems to be few  the stages from recon-all: https://surfer.nmr.mgh.harvard.edu/fswiki/recon-all (see section "Manual-Intervention Workflow Directives": stages 1-31)

\csvnames{csvcol}{1=\pipeline, 2=\nfunc, 3=\cputime}
\csvstyle{interpolation}{
	tabular = |r|c|c|,
	table head = \hline Pipeline & \# of functions & \% CPU Time \\
	& with interpolation & from interpolation\\\hline\hline,
	late after line = \\\hline,
	respect all,
	csvcol
}
\begin{table*}[ht]
	\centering
	\csvreader[interpolation]{tables/interpolation.csv}{}
	{\pipeline & \nfunc & \tablenum[round-precision=2]{\cputime}}
	\caption{Contribution of interpolation to the applications' total CPU time. The percentage is the average sum of CPU time of functions using interpolation. The data includes all functions; not only the top 80\% of the CPU time.}
	\label{extab:interpolation}
\end{table*}

\subsection{ANTS: Single vs. Double Precision}
Figure~\ref{fig:makespan-ants} shows the makespan of ANTS brainExtraction and ANTS registrationSyN in double and single precision. For both pipelines, the makespan is significantly higher in single precision. This is unexpected, as the floating-point arithmetic operations are supposed to be faster in single precision~\cite{Wang2018-jv}.

@@

The number of iteration between both version of ANTS registrationSyN is similar. In fact, for the latest two stages, they were the same across all but one execution. However, The single precision version of ANTS registrationSyN took longer per iteration than the double precision version (Figure~\ref{fig:mean-time-per-iteration-ants}). Therefore, the slowdown is not due to the convergence of the algorithm, but rather the execution time for each iteration. \TG{This is too inconclusive, you should give all the details here.}

@@

\subsection{FreeSurfer: Thread-Synchronization}
Figure~\ref{fig:hotspots-freesurfer-reconall} shows a significant difference in the pipeline profile between single-threaded and multi-threaded execution. In the multi-threaded executions (Figure~\ref{subfig:hotspots-freesurfer-reconall-32threads}), OMP is a major bottleneck for the application, accounting for at least 76\% of the application runtime. FreeSurfer uses the multi-processing API provided by OpenMP. Moreover, the y-axis scale is multi folds larger than for the single-threaded executions. This indicates that the multi-threaded execution is significantly slower than the single-threaded execution.

@@

\section{Discussion}

\TG{I think the discussion should be merged in the results. I know it's unconventional but I think it would make the paper easier to read and more interestingly. You could add a short conclusion that summarizes the main take-aways from the results.}

\subsection{Few functions account for most of the CPU time}
Our initial assumption was that the distribution of CPU time would follow a Pareto distribution. Thus, 20\% of the functions would contribute to 80\% of the total CPU time. However, Figure~\ref{fig:long-tail-distribution} shows \TG{Figure 2 doesn't ``show'' that. Rephrase.} that the distribution of CPU time is closer to a Zipf's law with exponent $\alpha=1.8$. This means that the CPU time of a function is approximately inversely proportional to its rank. This is consistent with the Pareto's principle, but is more extreme. Therefore, future efforts to optimize the main bottlenecks of MRI pre-processing pipelines could lead to enormous performance improvements.

\subsection{Wasted CPU cycles due to Memory Bound}
Overall, our results show a high percent of memory bound across all pipelines.
Optimizing the data access of the application could reduce the number of data load. This would potentially have a large impact on the performance of the pipelines. Alternatively, the use of reduced precision could reduce the memory footprint, thus the memory bound. To reduce the memory bound impact of fetching data from disk, it could be possible to convert and store the data to a lower precision before pre-processing. In both cases, the impact of reduced precision on the accuracy of the pipelines should be studied.
\TG{You could cite some of Valerie's works in this direction.}

\subsection{Importance of Linear Interpolation}
Our results show that linear interpolation the primary bottleneck for all pipelines, except FreeSurfer recon-all. While each interpolation have low computational cost, the large amount of operations make it a significant bottleneck.

There is an extensive literature on interpolation \TG{cite it}. Unfortunately, large amount of optimization techniques are not applicable to MRI pre-processing pipelines, because of the rotation during registration \TG{explain how interpolation is related to registration. In your presentation of fmriprep, explain what registration is and the role it plays in the pipelines.} \TG{this is a bit cryptic and I don't think you want to state that here. You may want to say that it would be interesting to further investigate these techniques in the future.} which requires dynamic kernels for interpolation. Moreover, the high-resolution of the data and iterative refinement of the registration require a large number of interpolations to be performed.

The authors in \cite{Canelhas2018-vs} review different linear interpolation methods ranging from poorly performing nearest-neighbor to classical trilinear interpolation. They review algorithm trade-off accuracy for a lower computational cost. We think it would be interesting for future work to explore these different alternatives to accelerate interpolation within the pipelines. However, special attention will be required to balance the accuracy and the performance of the algorithm.

Similarly, we propose future efforts to explore reduced- or mixed-precision techniques to accelerate interpolation. Optimizing the data format used for interpolation would reduce the computational cost, memory footprint, and storage requirements.

Improvements to interpolation in 3D MRI imaging would have a large impact. It would also benefit other fields of studies using interpolation for high-resolution imaging with iterative refinement, such as climate simulation, astronomical imaging, or 3D image reconstruction. \TG{You could summarize the previous 3 paragraphs, I find them a bit verbose.}

>>>>

\usepackage{cite}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=black,citecolor=blue,filecolor=black,urlcolor=blue}
\graphicspath{{figures}}
\usepackage{caption}
\usepackage{subcaption}
\captionsetup[subfigure]{singlelinecheck=false, skip=0pt}
\usepackage{listings}
\renewcommand{\lstlistingname}{Listing}
\usepackage[binary-units=true]{siunitx}
\usepackage{csvsimple}
\usepackage{booktabs}
\usepackage{amsmath}

@@

\begin{abstract}
	Magnetic Resonance Image (MRI) pre-processing is a critical step for neuroimaging analysis. However, the computational cost of MRI pre-processing pipelines is a major bottleneck for large cohort studies and some clinical applications. While High-Performance Computing (HPC) and, more recently, Deep Learning have been adopted to accelerate the computations, these techniques require costly hardware and are not accessible to all researchers. Therefore, it is critical to understand the performance bottlenecks of MRI pre-processing pipelines to improve their performance. Using Intel VTune profiler, we characterize the bottlenecks of several commonly used MRI-preprocessing pipelines from the ANTs, FSL, and FreeSurfer toolboxes. We found that few functions contribute to most of the CPU time, and that linear interpolation is the largest contributor. Data access is also  a substantial bottleneck. Last, we identified a bug in the ITK library that impact the performance of ANTs pipeline in single-precision and a potential issue with the OpenMP scaling in FreeSurfer recon-all. Our results provide a reference for future efforts to optimize MRI pre-processing pipelines.
\end{abstract}

@@

% - No comprehensive study of the performance bottleneck of MRI pre-processing.
\HL{To the best of our knowledge, this is the first study that} characterize the computational cost of several commonly adopted MRI pre-processing pipelines. The results of our analysis serve as a reference for future efforts to optimize MRI pre-processing pipelines.

@@

% - Pipelines
Neuroscientists developed various toolboxes to tackle the challenging task of pre-processing MRI data. We focus on the commonly accepted fMRIPrep~\cite{Esteban2019-bl} pipeline as it a comprehensive pre-processing application for both sMRI and fMRI. fMRIPrep combines several pipelines to produce a complete pre-processing pipeline. We also study its sub-pipeline separately to provide a finer grain analysis of the performance bottleneck.

% - VTune profiler
To profile the pipelines, we use the Intel VTune profiler~\cite{vtune-profiler}. The VTune profiler is a multi-language profiler with low performance overhead and provides information of functions at runtime level. We profile the pipeline using \HL{-- single-threaded and 32 threads to simplify the analysis and study the the multi-threading performance, repsectively}. We aggregate subjects' results for each pipeline to obtain an average performance profile. We compiled each pipeline with debug information to obtain human readable information on function and module names in VTune.

\section{Materials \& Methods}
\subsection{Pipelines}

fMRIPrep is a commonly used fMRI and sMRI pre-processing pipeline for neuroimaging, which is built on top of several well-known neuroimaging toolboxes such as ANTs, FSL, FreeSurfer and AFNI. Although the processing steps may vary based on the type of input data, the sMRI pipeline generally uses ANTs BrainExtraction~\cite{Tustison2010-gg,Avants2008-ea} to perform intensity corretion and skull-stripping of a T1 weighted image, FSL FAST~\cite{Zhang2001-hx} to segment the tissues of the extracted brain, and ANTs registrationSyN~\cite{Avants2008-ea} to register the segmented brain. When enabled, FreeSurfer recon-all~\cite{Dale1999-wu} is used for surface reconstruction. We mimic the fMRIPrep specific 3 \HL{phases reconall pre-processing, which (1) performs autorecon1 without skull-stripping, (2) imports an external skull-strip from ANTs brainExtraction, and (3) resumes reconstruction with autorecon2 and autorecon3}. The fMRI pipeline generally performs motion correction with FSL MCFLIRT~\cite{Jenkinson2002-od}, slice timing correction with AFNI 3dtshift~\cite{Cox1996-nk}, and co-registration with FSL FLIRT~\cite{Jenkinson2002-od,Jenkinson2001-eu,Greve2009-dw}.

We profile the entire fMRIPrep pipeline to obtain a coarse view of the pipeline's performance bottleneck. Additionlly, we profile the sub-pipelines of fMRIPrep to characterize the performance at a finer grain. While we used the same sub-pipeline and order as in fMRIPrep, we used the default parameters for most of the sub-pipelines which may vary from the default parameters of fMRIPrep. We omitted some sub-pipelines of fMRIPrep as they were not compatible with our dataset. For example, time-slicing correction was already performed.

\HL{ANTs brainExtraction and registrationSyN provides the user with the option to use a single or double precision version of the pipeline, leveraging the Insight Segmentation and Registration Toolkit (ITK)}~\cite{Yoo2002-ve}. \HL{We profile both version to understand the performance benefits of using reduced precision in the pipelines.}

\subsection{Data}
To ensure that our performance profiles are inclusive of different populations, we used data with a wide range of age and equal distribution of sex. We used the the OpenNeuro ds004513 v1.0.2 dataset~\cite{ds004513:1.0.2}. The anatomical, functional, and diffusion data from 20 health individuals was acquired from two different cohorts. The within-subject cohort has nine participants (mean age=43~yrs, std=7~yrs; 4 females) with two session: eyes open and eyes closed. The replication cohort has eleven participants (mean age=27~yrs, std=5~yrs; 6 females) with only the eyes open session.

\subsection{Profiling}
Profiling MRI pre-processing pipelines raises a few challenges. Neuroimaging pipelines use several programming languages including \textit{C}, \textit{C++}, \textit{Fortran}, \textit{Matlab}, and \textit{Python}, with several pipelines utilizing a combination of multiple languages. Pipelines are generally complex and computationally expensive.

The Intel VTune profiler addresses these challenges by offering multi-language support, low performance overhead, and information of functions at runtime level. However, common profiling challenges remain. Profilers require source code to be compiled with debug symbols to report interpretable information about the different functions and modules. Moreover, different input data can substantially impact the performance results due to conditional branching in the pipeline and convergence thresholds. We discuss our approach to these challenges in the next section.
\MD{TODO Cite effect from varying input data on numerical stability and reduced precision. Ask yohan for references}

First, to obtain human readable information on function and module names in VTune, we re-compiled each pipeline with debug information in a Docker image. For use on HPC systems, we created Apptainer~\cite{Kurtzer2017-bu} images using Docker2Singularity\footnote{\href{https://github.com/singularityhub/docker2singularity}{https://github.com/singularityhub/docker2singularity}} \lstlistingname~\ref{lst:vtune_example} illustrates the profiling of a pipeline by mounting the VTune binary to the Apptainer image during execution.

\begin{minipage}{\linewidth}
	\lstinputlisting[
		language={sh},
		caption={Profiling a pipeline within a Apptainer container, using VTune profiler.},
		captionpos=b,
		label=lst:vtune_example,
		numbers=left,
		frame=single,
		basicstyle=\footnotesize,
		numbersep=5pt,
		numberstyle=\tiny\color{gray},
		rulecolor=\color{black},
		tabsize=2,
	]{vtune-example.sh}
\end{minipage}

Last, we profile the pipeline using two threading configuration: single-threaded and 32 threads. The single-threaded approach simplifies the analysis and limits the potential overhead from I/O on the profiling. 
The 32-thread approach uses all available CPU cores from a compute node of our infrastructure to study the multi-threading performance of the pipeline. This is a more realistic usage for the user of the pipeline.

Before profiling each pipeline, we transfer the input data from the shared file system to the compute node. After profiling, we transfer back the output to the shared file system. This limits profiling variability due to I/O contingency.

The data generated by the VTune profiler is written to a shared file system, for administrative reasons. In principle, this could increase the profiling overhead. However, in practice there is no impact since the data generated by the VTune profiler is very small and is not written on the same file system as data produced by the pipelines.

\subsection{Metrics Definition}
VTune provides several metrics to characterize the performance of a pipeline.  The makespan is the elapsed time from the start to the end of the pipeline. The CPU time refers to the total time spent by the CPU, across all cores, to execute the pipeline.

The memory bound metrics counts the number of cycles where the CPU is starved while waiting for in-flight memory demand load. L1 bound occurs when a demand load stall some CPU cores although the data already reside in L1 cache. Similarly, L2, L3, and DRAM bound occurs when a demand load stall the CPU and there is a cache miss in the previous level of cache, but the data reside in the L2, L3, and DRAM respectively. The store bound occurs when a store operation stall the CPU. Figure~\ref{fig:memory-metrics} shows the monitoring events and equation to derive the metrics defining the memory bound characterization~\cite{Intel2006-lc}. We use the convention from~\cite{Kukunas2015-jd} for the equations.

@@

\HL{We aggregate the profiling results of each pipeline across all subjects to obtain an average performance profile. The Pareto principle is a power law which states that 80\% of the effects come from 20\% of the causes. We hypothesize that the Pareto principle applies to the performance of MRI pre-processing pipelines, which is consistent with the literature}~\cite{Kukunas2015-jd}. \HL{Therefore, for fine-grained analysis, we only study the functions with the largest contribution to CPU time, up to 80\% cumulative total runtime of the pipeline. This heuristic allows us to focus our efforts on the performance critical sections of a pipeline. We report the mean and standard deviation of the execution time for those functions.}

\subsection{CPU time distribution across functions was long tailed}
Figure~\ref{fig:long-tail-distribution} shows that 80\% of the total CPU time was spent in less than 1.5\% of the function across all pipelines. While consistent with the Pareto principle, the distribution of the functions' CPU time does not follow a Pareto distribution. Using the \textit{powerlaw} Python package~\cite{Alstott2014-gr} we found that the data fit a Zipf's law with exponent $\alpha=1.46$ (see Figure~\ref{fig:zips-law}). Therefore the CPU time of a function is approximately inversely proportional to its rank. This is consistent with the Pareto's principle, but is more extreme. \HL{In other words, most of the CPU time is spent in very few functions. By consequence, future efforts to optimize this minimal set of functions could result in significant performance improvement.}

@@

\subsection{Main Bottleneck: Linear Interpolation}
Table~\ref{extab:interpolation} shows that interpolation is a critical part for all pipelines except FreeSurfer recon-all, contributing between 32.20\% to 62.70\% of the total CPU time. Overall, the number of functions using interpolation is low with less than 20 in each pipelines. In other words, few interpolation functions contribute to a large portion of the CPU time. While each interpolation have low computational cost, the large amount of operations make it a significant bottleneck. Therefore, small optimization to the interpolation operation would bring substantial benefits to these pipelines.

\csvnames{csvcol}{1=\pipeline, 2=\nfunc, 3=\cputime}
\csvstyle{interpolation}{
	tabular = |r|c|c|,
	table head = \hline Pipeline & \# of functions & \% CPU Time \\
	& with interpolation & from interpolation\\\hline\hline,
	late after line = \\\hline,
	respect all,
	csvcol
}
\begin{table*}[ht]
	\centering
	\csvreader[interpolation]{tables/interpolation.csv}{}
	{\pipeline & \nfunc & \tablenum[round-precision=2]{\cputime}}
	\caption{Contribution of interpolation to the pipelines' total CPU time. The percentage is the average sum of CPU time of functions using interpolation. The data includes all functions; not only the top 80\% of the CPU time.}
	\label{extab:interpolation}
\end{table*}

Despite an extensive literature on linear interpolation \MD{TODO cite it}, no particular optimization techniques are available in the MRI pre-processing pipelines we profiled. We think it would be interesting to explore some simple optimization techniques. For example, the authors in \cite{Canelhas2018-vs} review different linear interpolation methods ranging from poorly performing nearest-neighbor to classical trilinear interpolation. We think it would be interesting for future work to explore these different alternatives to accelerate interpolation in the pipelines, while considering the trade-off between accuracy and computational cost. Reduced- or mixed-precision techniques could also be explored to accelerate interpolation, with similar trade-off. Optimizing the data format used for interpolation would reduce the computational cost, memory footprint, and storage requirements. Improving the performance of interpolation would have a large impact for 3D MRI imaging processing, and other fields such as climate simulation, astronomical imaging, or 3D image reconstruction.

\subsection{Impact of Memory Bounded Functions}
Table~\ref{extab:memory-bound} shows that fetching data from disk \HL{(Memory bound - sum of other metrics)} is the main source of cycles where the CPU is starved. Following, the L1 bound has the highest cache-level contribution to the memory bound of the pipelines. ANTS pipelines have the highest L1 bound, followed by FreeSurfer recon-all and FSL MCFLIRT. L2 and L3 bound is low for all pipeline, although highest for FreeSurfer recon-all. The DRAM bound is high for both FSL FLIRT and FreeSurfer recon-all, followed by ANTS brainExtraction in single and double precision. Store bound is low for all pipelines. We note an interesting relationship between the low L1 bound and the high DRAM bound of FreeSurfer recon-all and FSL FLIRT, compared to the other pipelines.

Optimizing the data access of the pipeline could reduce the number of data load. This would potentially have a large impact on the performance of the pipelines. Alternatively, the use of reduced precision could reduce the memory footprint, thus the memory bound. To reduce the memory bound impact of fetching data from disk, it could be possible to convert and store the data to a lower precision before pre-processing. In both cases, the impact of reduced precision on the accuracy of the pipelines should be studied.
\MD{TODO TG: You could cite some of Valerie's works in this direction.}

@@

\subsection{ANTs: Single vs. Double Precision}
Figure~\ref{fig:makespan-ants} shows the makespan of ANTS brainExtraction and ANTS registrationSyN in double and single precision. For both pipelines, the makespan is significantly higher in single precision. This is unexpected, as the floating-point arithmetic operations are supposed to be faster in single precision~\cite{Wang2018-jv}.

@@

ANTs registrationSyN is a three stage process (Euler transform, affine transform, and SyN registration), each having 4 level with a specified number of maximum iterations. We focus on the SyN registration stage since it is multiple order longer to excecute. The number of iterations between both version of ANTs registrationSyN is similar. In fact, for the last two levels, the maximum specified number of iterations is reached across all but one execution. However, the single precision version of ANTS registrationSyN took \HL{approximately 20-25\%} longer per iteration than the double precision version for the last registration stage (Figure~\ref{fig:mean-time-per-iteration-ants}). Therefore, the slowdown is not due to the convergence of the algorithm, but rather the execution time for each iteration.

After further analysis of the ANTs registrationSyN single-precision pipeline, we found that the template\footnote{\href{https://github.com/InsightSoftwareConsortium/ITK/blob/d9c585d96359bf304ad3047148cee81bf27ac0c1/Modules/Core/ImageFunction/include/itkVectorInterpolateImageFunction.h\#L46-L48}{itkVectorInterpolateImageFunction.h}} forces the output of its function to be in double precision, with it's \textit{EvaluateAtContinuousIndex} contributing most of the CPU time. Therefore, the pipeline wastes CPU cycles to convert the input data to double in ITK, and potentially to cast back to float in ANTS. We think this bug causes the slow-down. We tried recompiling a version of ITK to fix this issue, but we were unsuccessful due to the size and complexity of the code base. This example shows that the use of reduced precision is not trivial and require a deep understanding of the code base. Moreover, benchmark should be performed to ensure that the reduced precision does not impact the performance or accuracy of the pipeline.
We reported the issue on ITK GitHub repository at: \href{}{} \MD{TODO open issue}

@@

\subsection{FreeSurfer: Thread-Synchronization}
Figure~\ref{fig:hotspots-freesurfer-reconall} shows a significant difference in the pipeline profile between single-threaded and multi-threaded execution. In the multi-threaded executions (Figure~\ref{subfig:hotspots-freesurfer-reconall-32threads}), OMP is a major bottleneck for the pipeline, accounting for at least 76\% of the pipeline runtime. FreeSurfer uses the multi-processing API provided by OpenMP. Moreover, the y-axis scale is multi folds larger than for the single-threaded executions. This indicates that the multi-threaded execution is significantly slower than the single-threaded execution.

@@

OpenMP offers multiple scheduling policies to assign chunks to threads including -- \textit{static} scheduling assigns chunks to threads in round-robin fashion and \textit{dynamic} which assign chunks to threads as they become available for work. In the codebase 83 out of 92 tasks use the \textit{static} scheduling policy. This could lead to a load imbalance between the threads, and thus a decrease in parallel efficiency. We speculated that \textit{dynamic} scheduling could improve the parallel efficiency, by assigning chunks based on the current load of the threads. Unfortunately, we failed to recompile FreeSurfer by naively changing the \textit{static} OpenMP scheduling to \textit{dynamic}. This remains challenging as the codebase is large and complex, and the change in scheduling policy could lead to unexpected bugs.
We opened an issue on FreeSurfer GitHub repository at: \href{}{} \MD{TODO open issue}
